Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-03-01 analysis/paper/paper.Rmd
editing based on KC's comments

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the early Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The representative ceramic styles in the early Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, suggesting a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

<!--info about Gaoshan-->
To compare, we examine a sample from Gaoshan in the southwestern Chengdu Plain, an alluvial plain located between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013; @Gaoshan2017]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020]. It is suggested that the spread of rice farming into the Sichuan area occurred at least 4500 years ago from the Middle and Lower Yangzi valleys [@Chi2010].

Although the current findings at Guijiabao and Gaoshan support the spread of mixed agriculture into southwest China around 5000 BP, the proportions of millet or rice included in local culinary practices during the early Neolithic period remain unclear. The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in the neighboring sites, such as Gaoshan, Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013]. The approach of residue analysis for ceramics reflecting food preparation and processing could be helpful to understand the adoption of new practices.

@@

## Samples
We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan. The fragments were from known contexts with radiocarbon dates: early Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)). Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Hao2022, @Guijiabao2017a]. This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.

@@

The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at XXX to enable research reproducibility and transparency [@Marwick2017].

@@

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but in relatively small amounts in the majority of them (Table \@ref(tab:FA-composition-table)]. The lipid profiles show that the main compounds in most potsherds are saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from larger molecules, such as triacylglycerols, after long-term burial processes due to oxidation and microbial degradation [@Dudd1998; @Eerkens2007]. High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. Although a higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao abd Gaoshan. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

@@

To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

The results show that the majority of Guijiabao potsherd samples contain fatty acids from ruminant adipose fats, with one potsherd (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. The potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value, suggesting a C4 plant diet for the ruminant animal.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots. The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found a base sample dating to the early Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation. We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. Also, a detailed analysis of faunal remains from Guijiabao is still in progress, which constrains our discussion. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices in Guijiabao.

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (e.g., rice), C4 plants (e.g., millets), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. It is also possible that other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. However, the seeds of wild weeds only account for a small part of the plant remain. Surprisingly, we only observed signals from C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Freshwater fish could be another origin of non-ruminant fats based on the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], but we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses according to unearthed faunal remains from the sites, which were accompanied with stone tools for hunting, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a]. However, whether horse was a part of local food choices during the Neolithic period is uncertain. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the early Neolithic. The evidence from Guijiabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].

We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). It appears that C3 plants could be processed more in the historical period compared to the early Neolithic. But it needs more studies to explore whether there is a shift in diet over time. The overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important food source for local animals and could also be included in the local diet at Guijiabao and Gaoshan during the Neolithic transition. Despite no C4 plant origin being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants. The evidence of stone knives and tools also indicates the practice of agricultural harvesting [@Zhou2019; @Hao2022; @Lin2022]. We also note that our finding of a more C3-based diet based on ceramic residues is not consistent with the result of a C4-based diet according to isotopic analysis of human bones [@Lin2022]. The contradiction could be caused by different sampling strategies yielding contradictory results or by residue analysis reflecting only a portion of the content of food practices. Nevertheless, all pieces of evidence support the existing theory of mixed rice/millet farming in southwest China. However, farming practices may not have been adopted as much as we initially thought.

In general, during the Neolithic transition, the food source detected from our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals. Unearthed stone tools, including arrowheads and balls, also suggest hunting activities. Considering that Guijiabao is located in highland areas and Gaoshan is close to Hengduan mountain ranges, a subsistence relying on hunting to a certain degree could be expected. Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the early Neolithic, and it might only account for a small proportion. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison. An organic residue approach may only provide a narrow perspective on understanding local diets, but it could improve our current understanding of food sources through direct evidence cooked in ceramics. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in the local diet during the Neolithic transition.

>>>>

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, suggesting a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

<!--info about Gaoshan-->
To compare culinary practices for different ecological zones, we examine a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020]. It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from the Middle and Lower Yangzi valleys [@Zhang2010].

Although the current findings at Guijiabao and Gaoshan support mixed agriculture, the proportions of millet or rice included in local culinary practices during the Neolithic period remain unclear. The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice at most sites in southwest China, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017], and Henglanshan and Baiyangcun in Yunnan [@Hao2022]. The approach of residue analysis for ceramics reflecting food preparation and processing could be helpful to understand the adoption of new practices.

@@

## Samples
We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan. The fragments were from known contexts with radiocarbon dates: Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)). Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Hao2022, @Guijiabao2017a]. This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.

@@

The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017].

@@

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but in relatively small amounts in the majority of them (Table \@ref(tab:FA-composition-table)]. The lipid profiles show that the main compounds in most potsherds are saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from larger molecules, such as triacylglycerols, after long-term burial processes due to oxidation and microbial degradation [@Dudd1998; @Eerkens2007]. High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005]. However, we may still need to examine their isotopic values to confirm their sources.

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao and Gaoshan. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

The results show that the majority of Guijiabao potsherd samples contain fatty acids from ruminant adipose fats, with one potsherd (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaoshan (CDG-062) demonstrates ruminant adipose fats as well. The potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting those ruminant animals seemed to have consumed mainly C3 plants. There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value, suggesting a C4-based diet for the ruminant animal.

@@

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots. The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found a base sample dating to the early Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation. We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. Also, a detailed analysis for faunal remains from Guijiabao is still in progress, which constrains the comparison between different lines of evidence. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices in Guijiabao.

<!--discussion about possible plant origins-->
While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were dominated cereal according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples. Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. It is also possible that native wild weeds classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet, but those seeds only account for a small part of the plant remain according to flotation results [@Guijiabao2016]. Despite stone knives and tools were also found at the sites, it is uncertain if they were used for the practice of agricultural harvesting [@Zhou2019; @Hao2022; @Lin2022]. Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b]. However, we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses according to unearthed faunal remains from the sites, which were accompanied with stone tools for hunting, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a]. However, whether horse was a part of local food choices during the Neolithic period is uncertain. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China with only sporadic presence in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].

We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). It appears that C3 plants could be processed more in the historical period compared to the early Neolithic. The source of C3 plants might also include wheat since it was introduced in the historical period [@JiangGaoshan2017]. More studies and larger sample size are needed to explore whether there is a shift in diet over time. The overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important food source for local animals and could also be included in the local diet at Guijiabao during the Neolithic period. Despite no C4 plant signal being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants. We also note that that our finding of C3 plant signal based on ceramic residues is not consistent with the result of a C4-based diet according to isotopic analysis of human bones [@Lin2022]. The contradiction could be caused by different sampling strategies yielding contradictory results or by residue analysis reflecting only a portion of the content of food practices. Nevertheless, all pieces of evidence support the existing theory of mixed rice/millet farming in southwest China. Our results indicate that farming practices may not have been adopted as much as we initially thought.

In general, during the Neolithic period, the food source detected from our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer. Unearthed stone tools, including arrowheads and balls, also suggest hunting activities. Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges suggested by the case of Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected. Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, and it might only account for a small proportion. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison. An organic residue approach may only provide a narrow perspective on understanding local diets, but it could improve our current understanding of food sources through direct evidence cooked in ceramics. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in the local diet during the Neolithic period.
